Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-02-02 Multiresolution_framework/ltx/manuscript.tex
rearranged the algorithms

In summary, to solve the joint state and parameter estimation for linear state-space models one can adapt the EM algorithm, solving the smoothing problem for the E-step and constructing and maximising the lower bound in the M-step. A schematic representation of the EM algorithm is represented in Figure~\ref{fig:EMBlockDiagram}.  
\begin{figure}[!t]
\centering
\includegraphics[scale=1]{./Graph/EMBlockDiagram.pdf}
\caption{{\bf Schematic of the EM algorithm}. The RTS smoother (E-step) and the maximisation of the lower bound, $\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')$, (M-step) are shown. The dashed box shows the iterative two step algorithm once it is initialised.}
\label{fig:EMBlockDiagram}
\end{figure}
  \subsection{E-step}
  As mentioned earlier the E-step for state-space models is equivalent to the smoothing problem. We use the RTS smoother in order to infer the states of the system. The RTS smoother uses the Kalman filter \cite{Kalman1960} in forward pass along withe a backward iteration to calculate its outputs: smoothed states $\hat{\mathbf x}^b_t$ and covariances, $\mathbf P^b_t=\mathrm{cov}(\mathbf{x}_t)$. Following \cite{Gibsona2005}, an additional backwards recursion is also included (step 6 of the RTS algorithm)	to calculate cross-covariance matrix, $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t},\mathbf{x}_{t+1})$, which is also required in forming the lower bound to be maximised in the M-step. Note that the following quantities from the forward iteration (filtering) should be stored when implementing the RTS smoother: predicted and corrected estimates for both state estimates ($\mathbf{x}_t^{f-}$, $\mathbf{x}_t^{f}$) and covariance matrices ($\mathbf P_t^{f-}$,$\mathbf P_t^f$). Also the final value of the Kalman gain, $\mathcal K_T$, should be stored in order to initialise the backward cross-covariance computation. 
 \subsection{M-step}
It is rather straightforward to show that 
\begin{equation}
	\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')= \mathbf E_{\boldsymbol \theta'}\left[2\ln p(\mathbf X,\mathbf Y;\boldsymbol \theta)\right], 
\end{equation}
is a lower bound on the log-likelihood function \cite{Bishop2006} (see Appendix~\ref{ap:Lowerbound} )  where $p(\mathbf X,\mathbf Y;\boldsymbol \theta)$ is the joint probability distribution of the states and the observations and where $ \mathbf E_{\boldsymbol \theta'}\left[\cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, $\boldsymbol\theta'$ i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). Note that this is actually the smoother output.

@@

The algorithm has two steps: the E-step, which computes $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step, which updates the parameter estimates by computing the (analytic) maximum of $Q(\boldsymbol\theta,\boldsymbol\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge. The RTS  algorithm is given in Algorithm~\ref{alg:MRA-RTS} for completeness.

@@

\caption{\textbf{State and parameter estimation for the MRAIDE model.} This table shows the steps for joint state and parameter estimation using the EM based algorithm.}

>>>>

In summary, to solve the joint state and parameter estimation for linear state-space models one can adapt the EM algorithm, solving the smoothing problem for the E-step and constructing and maximising the lower bound in the M-step. A schematic representation of the EM algorithm is represented in Figure~\ref{fig:EMBlockDiagram}.  
\begin{figure}[!t]
\centering
\includegraphics[scale=1]{./Graph/EMBlockDiagram.pdf}
\caption{{\bf Schematic of the EM algorithm}. The RTS smoother (E-step) and the maximisation of the lower bound, $\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')$, (M-step) are shown. The dashed box shows the iterative two step algorithm once it is initialised.}
\label{fig:EMBlockDiagram}
\end{figure}
  \subsection{E-step}
  As mentioned earlier the E-step for state-space models is equivalent to the smoothing problem. We use the RTS smoother in order to infer the states of the system. The RTS smoother uses the Kalman filter \cite{Kalman1960} in forward pass along withe a backward iteration to calculate its outputs: smoothed states $\hat{\mathbf x}^b_t$ and covariances, $\mathbf P^b_t=\mathrm{cov}(\mathbf{x}_t)$. The RTS  algorithm is given in Table~\ref{alg:MRA-RTS}. Following \cite{Gibsona2005}, an additional backwards recursion is also included (Step~5 of the RTS algorithm)	to calculate cross-covariance matrix, $\mathbf M_t=\mathrm{cov}(\mathbf{x}_{t},\mathbf{x}_{t+1})$, which is also required in forming the lower bound to be maximised in the M-step. Note that the following quantities from the forward iteration (filtering) should be stored when implementing the RTS smoother: predicted and corrected estimates for both state estimates ($\mathbf{x}_t^{f-}$, $\mathbf{x}_t^{f}$) and covariance matrices ($\mathbf P_t^{f-}$,$\mathbf P_t^f$). Also the final value of the Kalman gain, $\mathcal K_T$, should be stored in order to initialise the backward cross-covariance computation. 
 \subsection{M-step}
It is rather straightforward to show that 
	\mathcal Q(\boldsymbol \theta,\boldsymbol\theta')= \mathbf E_{\boldsymbol \theta'}\left[2\ln p(\mathbf X,\mathbf Y;\boldsymbol \theta)\right], 
is a lower bound on the log-likelihood function \cite{Bishop2006} (see Appendix~\ref{ap:Lowerbound} )  where $p(\mathbf X,\mathbf Y;\boldsymbol \theta)$ is the joint probability distribution of the states and the observations and where $ \mathbf E_{\boldsymbol \theta'}\left[\cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, $\boldsymbol\theta'$ i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). Note that this is actually the smoother output.

@@

The algorithm has two steps: the E-step, which computes $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step, which updates the parameter estimates by computing the (analytic) maximum of $Q(\boldsymbol\theta,\boldsymbol\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge. A summary of the joint state parameter estimation is given in Table~\ref{alg:EMsteps}.

@@

\caption{\textbf{State and parameter estimation for the MRAIDE model.} This table shows the steps for the joint state and parameter estimation using the EM based algorithm.}
